Icicle: write once, run once

### 2016-06-02 2016fhpc/section/02-Source.tex
minor changes
start on core conversion of mean group

$\mi{Prim}$ defines the primitive operators.
Function and query definitions are given in $\mi{Def}$. $\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries. All queries in a top-level program process the same table.

@@

Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$. Rule TcBox assigns a pure value either @Element@ or @Aggregate@ type.

@@

The rules PrimArith, PrimRel and PrimLookup assign types to primitives. Rule AppArgs produces the type of a function or primitive applied to its arguments. Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.

@@

% ---------------------------------------------------------
\subsection{Evaluation}
The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.

Definition $V$ are simple values of numbers, booleans, maps and tuples.
Definition $V'$ are stream values that correspond to modalities: pure computations are represented as a single value;
@Element@s as a stream transformer from a heap to a value;
and @Aggregate@s as folds defining an initial state and update and extract functions.
Definition $M'$ are the modalities themselves without value types.

The evaluation rules operate over a slightly modified expression type, $E'$, which is an $\mi{Exp}$ with $V'$ as values instead of just simple values and natural numbers.
This is to simplify typechecking over $V'$, and as $V'$ only appear in restricted form for evaluation.
We also assume that all functions have been inlined before evaluation.

The judgment $\SourceStepX{M'}{E'}{V'}$ defines a big-step evaluation relation.
The rules below use the modality in deciding whether pure values should be converted to streams or folds.

Rule EVal applies when the expression is alread a stream value.
This corresponds roughly to the TcNat typechecking rule.

Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the environment modality is not pure.
The stream transformer ignores the input stream and returns the pure value.
The fold has a unit accumulator, and the extract function simply returns the pure value.

Rules EPrimValue, EPrimStream and EPrimFold evaluate primitive applications.
For pure values, the values are unboxed and the primitive is applied.
For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
Finally, the extract function extracts the final values of the input folds, and applies the primitive.

Rule ELet rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
The only difference is we inspect the type of the bound expression to find the modality, then evaluate the bound under that modality.
The rest of the expression is evaluated under the original modality.

Rule EFold is one of the most complex, as it introduces recursion to the streams.
It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
This sleight of hand is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
The extract function for this is the identity, simply returning the accumulator.

@@

Rule EGroup is rather similar to filter, except that it returns a map.
Here, we start with an empty map of bottoms.
At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
We then apply the update function to the accumulator, and update that key in the map.
For the extract function, we simply apply the extract to each element of the map.

In order to execute these evaluation rules on an actual input table, a few steps are necessary.
First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
For example, $e[@open@~:=~@Stream@~(\lam{\store} \store~@open@)]$.
Next, the query must be converted to a @Fold@.
Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
This occurs for every record in the table.
Finally, the extract function is called, which gives the result of the query.

>>>>

$\mi{Prim}$ defines the primitive operators.
$\mi{V}$ defines values.
$\mi{Def}$ contains both function and query definitions.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
All queries in a top-level program process the same table.

@@

Rules TcNat, TcBool, TcMap and TcPair assign types to literal values.
Rule TcVar performs variable lookup in the context.
Rule TcBox assigns a pure expression either @Element@ or @Aggregate@ type.

@@

The rules PrimArith, PrimRel, PrimLookup, PrimFst and PrimSnd assign types to primitives.
Rule AppArgs produces the type of a function or primitive applied to its arguments.
Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.

@@

% ---------------------------------------------------------
\subsection{Evaluation}
The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
Grammar $N$ defines the modes of evaluation, including pure computation.
Grammar $\Sigma$ defines a heap containing stream values.
Grammar $V'$ defines a different type of stream value for each evaluation mode:
\begin{itemize}
\item
@Pure@ computations are represented as a single value;
\item
@Element@ stream transformers are represented as a function applied to each element; and
\item
@Aggregate@ folds are represented as an initial (zero) state, an update (konstrukt) function applied to each stream element and current state, and an eject function applied to the final state.
\end{itemize}

The judgment form $\SourceStepX{N}{\Sigma}{e}{V'}$ defines a big-step evaluation relation: under evaluation mode $N$ with heap $\Sigma$, expression $e$ evaluates to stream value $V'$.
The evaluation mode is used to know when pure values should be converted to streams or folds.
We assume that all functions have been inlined into the expression before evaluation.

Rule EVal applies when the expression is already a stream value.
Rule EVar performs variable lookup in the heap.
Rule ELet uses the bound expression's type to find the evaluation mode, then evaluates the bound expression under that mode.
The bound expression's value is added to the heap, and the rest of the expression is evaluated in the original evaluation mode.

Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the evaluation mode is not pure.
Conversion to a stream transformer ignores the input stream and returns the pure value.
Conversion to a fold has no state, and the eject function returns the pure value.

Rules EPrimValue, EPrimStream and EPrimFold evaluate primitive applications.
For pure values, the values are unboxed and the primitive is applied.
For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
For folds, the states of all the argument folds are joined together as tuples.
At each step, the states of the input folds are updated using their own update function.
The eject function extracts the ejected values of the input folds, and applies the primitive.

Rule EFold introduces recursion and memory to the streams.
It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
The update expression has mode @Element@, so it must be evaluated to a stream.
This stream expects the stream element as an argument, but we need to give it the current fold state as well.
In order to pass both as a pair, we modify all streams in the heap to use the second half of the pair.
We also add a stream for the fold state to the heap; this uses the first half of the pair.
The update function then calls the stream transformer with a pair of the state and the input element.
The eject function is the identity function.

@@

Rule EGroup is rather similar to filter, except that it returns a map.
Here, we start with a map of every key to the fold initialiser.
At every update step, we find the key and look up the current accumulator for that key in the map.
We then apply the update function to the accumulator, and update that key in the map.
For the eject function, we simply apply the eject to each element of the map.

The judgment form $t~|~e~\Downarrow~V$ evaluates an expression over a table input: on input table $t$, aggregate expression $e$ evaluates to value $V$.
The input table $t$ is a map from column name to a list of all the values for that column.
Rule ETable creates an initial heap where each column is a stream transformer that pulls out the $n$th element of a nested tuple.
Tuples are constructed correspondingly, so that each row becomes a nested tuple where the $n$th element contains the value of the $n$th column.

### 2016-06-02 2016fhpc/section/03-Core.tex
start on core conversion of mean group

%!TEX root = ../Main.tex
\section{Intermediate language}
\label{s:IcicleCore}

@@

\begin{tabbing}
MMM \= MM \= MMMM \= MM \= MMMMMM \= \kill
$\mi{PlanX}$
\GrammarDef
  $x~|~@Int@~|~\mi{Prim}~\ov{\mi{PlanX}}$
\\
\\
$\mi{Plan}$
\GrammarDef
  $@plan@~x$ \> $\{~\ov{x~:~T;}~\}$
\\
  \> \> $@before@$ \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
  \> \> $@folds@$  \> $\{~\ov{x~:~T~=~\mi{PlanX}~@then@~\mi{PlanX};}~\}$ \\
  \> \> $@after@$  \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
  \> \> $@return@$ \> $\{~\ov{x~:~T~=~x;}~\}$ \\
\end{tabbing}

\caption{Grammar}
\label{fig:core:grammar}
\end{figure}

@@

The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
Expressions can be variables or primitive applications: function applications are not allowed in expressions here, as their definitions are inlined before converting to query plans.
The @Plan@ itself is split into stages of computation, which correspond to the modal stages in the Source query.
The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
The next stage, @folds@, defines the @Element@ computations and how they are converted into @Aggregate@ computations.
Next are @after@ which correspond to @Aggregate@ computations, occurring after the entire stream has been seen.
Finally, the @return@ specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.

One of the simplest queries to convert to a query plan is a count.
The query plan header specifies the table and the type of each column in the table.
This defines a single fold called @count@ which starts at zero, and is incremented for each element.
At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
\begin{code}
    table stocks { open : Int; close : Int; }
    query count_query = count

==> plan stocks { open : Int; close : Int; }
    before { }
    folds  { count : Int = 0 then count + 1 }
    after  { }
    return { count_query : Int = count }
\end{code}

Computing the sum is only slightly more complicated.
Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
\begin{code}
    query sum_open = sum open

==> plan stocks { open : Int; close : Int; }
    before { }
    folds  { sum : Int = 0 then sum + open }
    after  { }
    return { sum_open : Int = sum }
\end{code}

In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
Here we use two folds, one for the sum, and one for the count.
While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
\begin{code}
    query mean_open = sum open / count

==> plan stocks { open : Int; close : Int; }
    before { }
    folds  { sum       : Int = 0 then sum + open 
             count     : Int = 0 then count + 1 }
    after  { mean      : Int = sum / count }
    return { mean_open : Int = mean }
\end{code}

If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.
\begin{code}
    plan stocks { open : Int; close : Int; }
    before { }
    folds  { c1 : Int = 0 then c1 + 1
             c2 : Int = 0 then c2 + 1
             s1 : Int = 0 then s1 + open 
             s2 : Int = 0 then s2 + open }
    after  { m1 : Int = s2 / c2 }
    return { count_query : Int = c1
             sum_open    : Int = s1
             mean_open   : Int = m1 }
\end{code}

The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
We can perform fairly standard common subexpression elimination to remove these.
\begin{code}
    plan stocks { open : Int; close : Int; }
    before { }
    folds  { c1 : Int = 0 then c1 + 1
             s1 : Int = 0 then s1 + open }
    after  { m1 : Int = s1 / c1 }
    return { count_query : Int = c1
             sum_open    : Int = s1
             mean_open   : Int = m1 }
\end{code}

The interesting thing here is just how simple fusion is.
Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
The single-pass restriction means that any source query can be expressed as a single fold over the data.
By keeping the folds separate we have exposed more opportunities for common subexpression elimination.

>>>>

%!TEX root = ../Main.tex
\clearpage
\section{Intermediate language}
\label{s:IcicleCore}
\TODO{PAGE LAYOUT REMOVE CLEARPAGE}

@@

\begin{tabbing}
MMM \= MM \= MMMM \= MM \= MMMMMM \= \kill
$\mi{PlanX}$
\GrammarDef
  $x~|~V~|~\mi{PlanP}~\ov{\mi{PlanX}}~|~\lam{x}\mi{PlanX}$
\\
$\mi{PlanP}$
\GrammarDef
  $\mi{Prim}~|~@mapUpdate@~|~@mapEmpty@~|~@mapMap@~|~@mapZip@$
\\
\\
$\mi{Plan}$
\GrammarDef
  $@plan@~x$ \> $\{~\ov{x~:~T;}~\}$
\\
  \> \> $@before@$ \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
  \> \> $@folds@$  \> $\{~\ov{x~:~T~=~\mi{PlanX}~@then@~\mi{PlanX};}~\}$ \\
  \> \> $@after@$  \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
  \> \> $@return@$ \> $\{~\ov{x~:~T~=~x;}~\}$ \\
\end{tabbing}

\caption{Query Plan Grammar}
\label{fig:core:grammar}
\end{figure}

@@

The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
Expressions can be variables, values, applications of primitives or anonymous functions.
Function definitions and uses are not allowed in expressions here, as their definitions are inlined before converting to query plans.
Anonymous functions are only allowed as arguments to primitives: they cannot be applied or stored in variables.
The $\mi{Plan}$ itself is split into stages of computation, which correspond to the modal stages in the Source query.
The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
The next stage, @folds@, defines the @Element@ computations and how they are converted into @Aggregate@ computations.
Next are @after@ which correspond to @Aggregate@ computations, occurring after the entire stream has been seen.
Finally, the @return@ specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.

We will start with a small example by converting just the @last key@ part of the group-mean query.
Here, we have the function definition for @last k@, which is defined as a fold with an initial value to describe a missing date @NO_DATE@, and the update value is the current @k@.
The current fold value is bound to @l@ but is not used.

\begin{code}
    table kvs { key : Date; value : Real }

    function last (d : Element Date)
     = fold l = NO_DATE then d;

    query last_key = last key
\end{code}

We now inline function definitions before converting to a query plan, renaming as necessary.
This is guaranteed to terminate because no recursion is allowed in function definitions.

\begin{code}
==> table kvs { key : Date; value : Real } 
    query last_key = (fold c = NO_DATE then key)
\end{code}

We can now convert this to a query plan.

\begin{code}
==> plan kvs { key : Date; value : Real; }
    before { }
    folds  { c : Date = NO_DATE then key }
    after  { }
    return { last_key : Date = c }
\end{code}

We will now convert the mean group.
Here, we have function definitions for @count@, @sum@ and @mean@.
Count is defined by starting from zero, and incrementing by one for each element.
Sum starts at zero and increments its argument.
Mean is then defined as sum divided by the count.

\begin{code}
    table kvs { key : Date; value : Real }

    function count
     = fold c = 0 then c + 1;

    function sum (e : Element Real)
     = fold s = 0 then s + e;

    function mean (e : Element Real)
     = sum e / count;

    query mean_all = group key of mean value
\end{code}

After function inlining, the result is this:
\begin{code}
==> table kvs { key : Date; value : Real } 
    query mean_all = group key
     of (fold foldS = 0 then foldS + value)
      / (fold foldC = 0 then foldC + 1)
\end{code}

Finally, we have the query plan.
Each fold inside the group becomes its own map which is initialised to empty.
At each iteration the fold update is applied to the value in the map, or the fold's initial value if the key is not yet present.
After all iterations are finished, the two maps are joined together and the pairs are divided.
\begin{code}
==> plan kvs { key : Date; value : Real; }

    before { }

    folds  { groupS   : Map Date Real
                = mapEmpty
             then mapUpdate groupS key 0
                  (\foldS. foldS + value)

           ; groupC   : Map Date Real
                = mapEmpty
             then mapUpdate groupC key 0
                  (\foldC. foldC + 1) }

    after  { groupM   : Map Date Real
                = mapMap (/) (mapZip groupS groupC) }

    return { mean_all : Map Date Real = groupM }
\end{code}

It is possible to convert this as a map of pairs of the sum and count.
Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
For example, a query @group key of count@ can reuse the already constructed @groupC@ map.

If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.

The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
We can perform common subexpression elimination to remove these.

The interesting thing here is just how simple fusion is.
Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
The single-pass restriction means that any source query can be expressed as a single fold over the data.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
start on core conversion of mean group

\TODO{update to talk about single CPU number.
Single CPU version 30 times faster than R, and in the same ballpark as @wc@ and @grep@.
Because the memory usage is bounded, we have a better idea of how many we can run on a single machine without running out of memory and stepping on each other's toes.
This means we can scale up to get quite close to the disk speed.
This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.}

>>>>

\TODO{update to talk about single CPU number.
Single CPU version 30 times faster than R, and in the same ballpark as @wc@ and @grep@.
Because the memory usage is bounded, we have a better idea of how many we can run on a single machine without running out of memory and stepping on each other's toes.
This means we can scale up to get quite close to the disk speed.
This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.
Blah~\cite{mcsherry2015scalability}
}

### 2016-06-02 2016fhpc/section/XX-Related.tex
start on core conversion of mean group

The most closely related transforms are induction variable elimination\cite{shivers1988control} and global value numbering\cite{rosen1988global}.
Induction variable elimination finds and removes accumulators that are linear functions of the iteration number.
This does not deal with non-linear functions such as sum, or mutually recursive accumulators.

Global value numbering works on an SSA form and is much more general than induction variable elimination.
Early versions such as Rosen et al\cite{rosen1988global} only worked on extended basic blocks with backedges removed.
Alpern et al\cite{alpern1988detecting} introduced cases for specific loop backedges, where a fixpoint is performed to find the congruence sets of loops.
More recently, Gulwani and Necula\cite{gulwani2004polynomial} improved upon this, supporting removal of more expressions, and imposing a polynomial time bound.
Global value numbering for loops can perform all optimisations here.

Common subexpression elimination is very closely related, but most imperative compilers only perform common subexpression elimination on straight-line computations with no control flow\cite{debray1992compiler}, or on loop invariant expressions\cite{bodik1998complete}.
Common subexpression elimination for functional languages would be able remove lone expressions, but does not deal with mutually recursive folds.

Temporal common subexpression elimination in Single Assignment C
allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
This is a more general case of the elimination opportunities that arise from loop unrolling.

For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
\begin{code}
int[] A;
int a = 1;
int b = 0;
while(int i = 0; i != size - 1; i++)  {
  int ax = f(A[i]);
  int bx = f(A[i+1]);
  a += ax;
  b *= bx;
}
\end{code}

I cannot think of a better example than this right now.

@@

Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.

The polyhedral model\cite{benabderrahmane2010polyhedral} analyses data dependencies in loops.
For a given iteration, the polyhedral model computes the iterations which must be executed before the given iteration.
By knowing the dependent iterations, one can restructure the loop so that the iterations are executed in a different order, while still performing any dependencies in order.
For example, a loop iterating over @i@, with loop body @A[i] = A[i-10]@ depends on the tenth previous iteration.
This loop could be executed in an order like @0, 1, 2,...@, but it could also stride by ten: @0, 10, 20..., 1, 11, 21,...@.
This allows fusion of any two loops if the intersection of the two loops' iteration spaces is not empty.
In our case the iteration order is fixed, as the program cannot control the order in which stream elements are received.
The polyhedral model may still be applicable to streaming computations where one could introduce certain-sized buffers, then process the buffer in parallel.
In this case the polyhedral model may expose more opportunities for duplicate elimination, but does not itself remove duplicates.

>>>>

The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}. These systems perform a clock, or \emph{rate} analysis of the streams being processed. Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.

\TODO{fill}
Advances in dataflow programming languages~\cite{johnston2004advances}.
Survey of stream processing~\cite{stephens1997survey}.
How to architect a query compiler~\cite{shaikhha2016architect}.
StreamIt~\cite{thies2002streamit}.

Kahn process networks are a flexible alternative to MapReduce~\cite{vrba2009kahn}.
Scheduling dynamic dataflow~\cite{buck1993scheduling}.
Co-iterative characterization~\cite{caspi1998co}.

Compare to short-cut fusion~\cite{coutts2007stream}.

@@

Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.
